Parallel Multi-objective optimization using co-evolutionary islands and search space overlapping

### 2016-06-10 hpmoon-hpca.tex
First draft

\author{Pablo Garc\'ia-S\'anchez, Julio Ortega, Jes\'us Gonz\'alez, Pedro Angel Castillo and Juan Juli\'an Merelo\affilnum{1}}

@@

\begin{abstract}
This paper...
\end{abstract}

@@

In \citep{Garcia16hpmoon} we tried to demonstrate that using overlapped sections of the chromosome in a coevolutionary multi-objective island-based algorithm can improve the quality of the solutions in the same computing time when the number of islands increases. We discovered that the most adequate overlapping size chosen varies depending on a relation between the number of islands and the indidivual size. Moreover, the experiments were conducted in a mono-processor and synchronous island model. This motivates us to continue this research line, but this time, proposing a new method to automatically adapt the size of the overlapped sections, but also, performing all the experiments in a real cluster with up to 128 nodes, with extra number of islands.

The hypothesis of this paper is that using overlapped sections of the chromosome, calculated depending the number of islands to use, in a coevolutionary multi-objective island-based algorithm can improve the quality of the solutions in the same computing time with respect to disjoint and partly overlaped methods. To demonstrate this, a new overlapping islands scheme has been compared with a baseline version of a distributed NSGA-II \citep{Deb00NSGAII}, a coevolutionary disjoint section approach, and a fixed-size overlapped method. Different benchmarks problems and different large number of islands have been used. Results show that this new technique can improve different quality indicators in the same amount of time when the problem size is large.

@@

To solve this issue, some authors have proposed the usage of Master-Slave approaches. For example, Durillo et al. \citep{Durillo08masterslave} compared different master-slave approaches TODO DESCRIBIR MEJOR ESTE TRABAJO. The method proposed by Hiroyasu \citep{Hiroyasu07discussion} generates offspring depending on the computation power. TODO Y ESTE

The work by Deb in 2003 was one of the first approaches for distributed MOEAs (dMOEAs). In that work, the dominance of solutions was divided in the islands using a transformation of coordinates. Their authors concluded that dividing the search space is a good idea, but achieve this is not trivial. Other approaches for dividing the search space include objective separation in different processors \citep{Xiao03specialized}, or divide in two populations: one for elite and other search \citep{Wang09parallel}. Martens, in \citep{Martens13asynchronous} generated a Barabasi-Albert network as the island topology, and selection based on migrating individuals from a not-crowded area and acceptance based on diversity.

More similar approaches to the one presented here were studied in the
next two works, also using cooperative coevolution for high-dimensional problems. Dorronsoro et al. in
\citep{Dorronsoro13superlinear} obtained super-linear performance
in several instances using cooperative coevolution, focusing each island on a portion of the chromosome. However, in some instances the parallel version 
improved the sequential version.  Kimovski et al. \citep{Kimovski15Parallel} proposed a
master-slave method that splits the population into several processors,
each one running in parallel a MOEA that only affects some portion of
the individuals. After a certain number of generations, the master
process receives all the sub-populations to be combined. Different
combination alternatives in this step are compared. Up to 8 processors
were used in the experimental setup. Our approach in this paper does not
broadcast all solutions to all islands for recombination, as previous
works, but only one solution to a random island, needing less
communication time. Besides, the maximum number of islands of
Dorronsoro or Kimovsky approaches was 8, while in this paper we have
used up to 128 islands.

@@

\begin{itemize}
\item Hypervolume (HV): measures the area formed by all non-dominated solutions found with, respect to a reference point. Higher values imply better quality of the PF.
\item Inverted Generational distance (IGD): calculates the distance of the obtained set of solutions to the optimal PF. Therefore, this metric requires the optimal PF found in the literature, or the theoretical one. In this metric, the lower the better. %Pablo: theoretical?
\item Spread (S): Measures the spread between solutions, taking into account the euclidean distance between consecutive solutions. As in previous metric, the lower value, the better, as it implies solutions distributed along all the PF.
\end{itemize}

@@

\subsubsection{Coevolution with disjoint islands} 
In this approach, each individual of size $L$ is split into $P$ chunks of size $L/P$. Every island $p$ only performs crossover and mutation on the $p_{th}$ part of the individuals. Figure \ref{fig:disjoint} describes this approach.

\begin{figure*}
\includegraphics[width=12cm]{islandDisjoint.jpg}
\caption{Disjoint algorithm: every island $p$ only modifies the $p_{th}$ components (in grey) of the individuals.}
\label{fig:disjoint}
\end{figure*}

\subsubsection{Coevolution with overlapping islands}
This approach is similar to the previous one, but every island also uses the $p+1$ and $p-1$ (module size) chunks of the individual for crossover and migration. Therefore, some kind of overlapping of the crossed and mutated parts exists between islands. Figure \ref{fig:overlapping} shows the affected parts of the individuals in
each island.

\begin{figure*}
\includegraphics[width=12cm]{islandNoDisjoint.jpg}
\caption{Overlapping algorithm: every island $p$ modifies the  $p+1$,
  $p_{th}$ and $p-1$  components (in grey) of the individuals.}
  \label{fig:overlapping}
\end{figure*}

\subsubsection{Adaptive coevolution with overlapping islands} 
As in the previous method, the sections to address are overlapped, but instead using one extra section on each side of the $p$ section ($p+1$ and $p-1$), it uses $c$ fragments on each side ($p+c$ y $p-c$), being $c$ a value depending on the number of islands. The results shown in \citep{Garcia16evostar} have been used as a base to calculate $c$. In that work, the overlapped method ($c=1$) obtained better results when the number of islands were >8. Therefore, we have used the formula $c=round(0.2*P-1)/2$ to calculate the extra sections to overlap. %TODO TERMINA Y PONER MAS EJEMPLOS Y DECIR QUE ESTA ES UNA PRIMERA APROXIMACION

% --------------------------------------------------------------

@@

The four approaches have been run with two different chromosome lengths ($L$): 512 and 2048. Different number of islands ($P$) have also been compared: 8, 32 and 128. This maximum number of islands have also been  used in previous work in the literature \citep{Martens13asynchronous}. The crossover and mutation chosen, SBX and polynomial, have been used previously by other authors in \citep{Durillo08masterslave}.

@@

\begin{table*}
\begin{center}
\begin{tabular}{|c|c|}
\hline
{\em Parameter Name} & {\em Value} \\ \hline
Number of islands ($P$) & 8, 32 and 128 \\ \hline
Chromosome size ($L$) & 512 and 2048 \\ \hline
Execution time (s) & 25 (for 512) and 100 (for 2048) \\ \hline \hline
Global population size ($N$) & 1024 \\ \hline
Selection type & Binary Tournament Selection \\ \hline
Replacement type & Generational \\ \hline 
Crossover type & SBX \\ \hline
Mutation  type & Polynomial\\ \hline
Mutation probability & 1/$L$ \\ \hline
Generations between migration & 5 \\ \hline
Individuals per migration & 1 \\ \hline
Selection for migration & Binary Tournament\\ \hline
Runs per configuration & 30 \\ \hline
\end{tabular}
\caption{Parameters and operators used in the experiments.}
\label{tab:parameters}
\end{center}
\end{table*}

The ECJ framework \citep{ECJ} has been used to run the
experiments. Specific operators have been developed as new modules for
ECJ, and they can be downloaded from our GitHub repository under a
LGPL V3 License
\footnote{\url{https://https://github.com/hpmoon/hpmoon-islands}}. The
island model has been executed synchronously, using the ECJ Internal
Island Model, in a CentOS 5.4 machine with Intel(R) Xeon(R) CPU E5520
@2.27GHz, 16 GB RAM, with Java Version 1.6.0\_16. As in this paper we
only focus on the behaviour of our model in a single machine scenario
(the islands share the same memory and processor), the migration time
will not be as important factor as in a real distributed system. This
task will be addressed in future works. %Pablo: no sé si decir esto
                                %aquí% 
%
%

@@

Finally, as happened in \citep{Garcia16hpmoon}, results show that all the quality indicators lowers when the number of islands is increased, as the subpopulations are smaller. This is consistent with the claim by Durillo in \citep{Dorronsoro13superlinear}, who found that cooperative coevolutionary MOEAs work better on bigger populations (more than 100 individuals).

@@

%High-performance problems that require a large number of decision
%variables can leverage the division of the decision space that parallel
%and distributed algorithms imply. This can be done in dMOEAs by dividing the
%chromosome into different parts, each one modified by a different
%island. This paper compares a baseline distributed NSGA-II with two
%different strategies to separate the chromosome (disjoint or overlapping
%parts), using a high number of islands. Results show that these 
%methods can achieve better quality metrics
%than the baseline in the same amount of time. This can be explained 
%because of the time reduction in crossover and mutation in the chromosomes,
%producing a higher number of generations and more modifications of the 
%solutions of the PF in each sub-population, and therefore improving
%quality of the global PF.%

%Results also show that when increasing the number of islands, the overlapping 
%method significantly improves the results with respect to the disjoint method.
%Therefore, the relation between the number of islands used, the global population size,
%and the length of the chromosome may be a key factor to decide if using the 
%overlapping method or the disjoint one. Studying this factor with more types 
%of problems, and new configurations of population size and chromosome lengths will be addressed in the future.%
%

%Also, distributed implementations in several systems (such as a GRID or cluster) with more different
%quantities of islands/processors, will be used to perform a scalability study of the different
%methods, being the transmission time between islands a relevant issue to
%address. New benchmarks and real problems will also be  used to validate
%our approach.

@@

TIN2014-56494-C4-3-P y TIN2015-67020-P (Ministerio de Economía y Competitividad), PROY-PP2015-06 (Plan Propio 2015 UGR),  y MSTR (PRY142/14, Fundación Pública Andaluza Centro de Estudios Andaluces en la IX Convocatoria de Proyectos de Investigación).

>>>>

\author{Pablo Garc\'ia-S\'anchez, Julio Ortega, Jes\'us Gonz\'alez, \\Pedro Angel Castillo and Juan Juli\'an Merelo\affilnum{1}}

@@

\begin{abstract}
Multi-objective optimization methods have gained interest in the last years because the performance they can achieve solving real-world problems. These kind of problems usually require a high-dimensional decision space, and multi-objective Evolutionary Algorithms have been successfully used in the past. Due to the large computational cost these algorithms require, different parallelization and distribution methods have been proposed, with different parallel and distributed computer architectures to cope with them. One of the most extended methods are the distributed co-evolutionary evolutionary algorithms, where each processor only deals with a part of the decision space, that is, the individuals belonging to the same sub-population (island) explore the same subset of decision variables.

In this paper, a new method to automatically adapt the overlapping sections of the individuals have been compared to other collaborative evolutionary methods.
Different number of islands, chromosome sizes and problems have been used. The analysis of the obtained experimental results, by using different 
metrics, shows that coevolutionary approaches can provide statistically significant improvements with respect to the base algorithm, being the method to automatically adapt the overlapping size the one to obtain better performance during the same execution time. Results also show that the relation of the number of 
islands (subpopulations) to the length of the chromosome (number of decision variables) is also a relevant factor to determine the most efficient alternative to distribute the decision variables.
\end{abstract}

@@

In our preliminary work in \citep{Garcia16hpmoon} we tried to demonstrate that using overlapped sections of the chromosome in a coevolutionary multi-objective island-based algorithm can improve the quality of the solutions in the same computing time when the number of islands increases. We discovered that the most adequate overlapping size may vary depending on a relation between the number of islands and the individual size. Moreover, the experiments were conducted in a mono-processor and synchronous island model. This motivates us to continue this research line, but this time, proposing a new method to automatically adapt the size of the overlapped sections, but also, performing all the experiments in a real cluster with up to 128 nodes, with extra number of islands.

The hypothesis of this paper is that using overlapped sections of the chromosome, calculated depending the number of islands to use, in a coevolutionary multi-objective island-based algorithm can improve the quality of the solutions in the same computing time with respect to the baseline, disjoint and partly overlapped methods. To demonstrate this, a new overlapping islands scheme has been compared with a baseline version of a distributed NSGA-II \citep{Deb00NSGAII}, a coevolutionary disjoint section approach, and a fixed-size overlapped method. Different benchmarks problems and different large number of islands have been used. Results show that this new technique can improve different quality indicators in the same amount of time when the problem size is large.

@@

To solve this issue, some authors have proposed the usage of Master-Slave approaches. For example,  \cite{Durillo08masterslave} compared different master-slave approaches TODO DESCRIBIR MEJOR ESTE TRABAJO. The method proposed by Hiroyasu \citep{Hiroyasu07discussion} generates offspring depending on the computation power. TODO Y ESTE

The work by \cite{Deb03distributed} was one of the first approaches for distributed MOEAs (dMOEAs). In that work, the dominance of solutions was divided in the islands using a transformation of coordinates. Their authors concluded that dividing the search space is a good idea, but achieve this is not trivial. Other approaches for dividing the search space include objective separation in different processors \citep{Xiao03specialized}, or divide in two populations: one for elite and other search \citep{Wang09parallel}. Martens, in \citep{Martens13asynchronous} generated a Barabasi-Albert network as the island topology, and selection based on migrating individuals from a not-crowded area and acceptance based on diversity.

More similar approaches to the one presented here were studied in the
next two works, also using cooperative coevolution for high-dimensional problems. 
\cite{Dorronsoro13superlinear} obtained super-linear performance
in several instances using cooperative coevolution, focusing each island on a portion of the chromosome. However, in some instances the parallel version 
improved the sequential version.  \cite{Kimovski15Parallel} proposed a
master-slave method that splits the population into several processors,
each one running in parallel a MOEA that only affects some portion of
the individuals. After a certain number of generations, the master
process receives all the sub-populations to be combined. Different
combination alternatives in this step are compared. Up to 8 processors
were used in the experimental setup. Our approach in this paper does not
broadcast all solutions to all islands for recombination, as previous
works, but only one solution to a random island, needing less
communication time. Besides, the maximum number of islands of
Dorronsoro or Kimovsky approaches was 8, while in this paper we have
used up to 128 islands.

@@

\begin{itemize}
\item Hypervolume (HV): measures the area formed by all non-dominated solutions found with, respect to a reference point. Higher values imply better quality of the PF. Figure \ref{fig:hypervolume} clarifies how the hypervolume is calculated.
\item Inverted Generational distance (IGD): calculates the distance of the obtained set of solutions to the optimal PF. Therefore, this metric requires the optimal PF found in the literature, or the theoretical one. In this metric, the lower the better. %Pablo: theoretical?
\item Spread (S): Measures the spread between solutions, taking into account the euclidean distance between consecutive solutions. As in previous metric, the lower value, the better, as it implies solutions distributed along all the PF.
\end{itemize}

\begin{figure}
\centering
\includegraphics[width=6cm]{hypervolume.pdf}
\caption{Hypervolume calculation. Areas from Pareto Front solutions (black points) to a reference point (white point) are added.}
\label{fig:hypervolume}
\end{figure}

@@

\subsubsection{Coevolution with disjoint islands (D)} 
In this approach, each individual of size $L$ is split into $P$ chunks of size $L/P$. Every island $p$ only performs crossover and mutation on the $p_{th}$ part of the individuals. Figure \ref{fig:disjoint} describes this approach.

\begin{figure*}
\centering
\includegraphics[width=12cm]{islandDisjoint.jpg}
\caption{Disjoint algorithm: every island $p$ only modifies the $p_{th}$ components (in grey) of the individuals.}
\label{fig:disjoint}
\end{figure*}

\subsubsection{Coevolution with overlapping islands (O)}
This approach is similar to the previous one, but every island also uses the $p+1$ and $p-1$ (module size) chunks of the individual for crossover and migration. Therefore, some kind of overlapping of the crossed and mutated parts exists between islands. Figure \ref{fig:overlapping} shows the affected parts of the individuals in
each island.

\begin{figure*}
\centering
\includegraphics[width=12cm]{islandNoDisjoint.jpg}
\caption{Overlapping algorithm: every island $p$ modifies the  $p+1$,
  $p_{th}$ and $p-1$  components (in grey) of the individuals.}
  \label{fig:overlapping}
\end{figure*}

\subsubsection{Adaptive coevolution with overlapping islands (A)} 
As in the previous method, the sections to address are overlapped, but instead using one extra section on each side of the $p$ section ($p+1$ and $p-1$), it uses $c$ fragments on each side ($p+c$ and $p-c$), being $c$ a value depending on the number of islands. As a first approach to automatically calculate this value the results shown in \citep{Garcia16hpmoon} have been used as a base to obtain $c$. In that work, the overlapped method ($c=1$) obtained better results when the number of islands were >8. Therefore, we have used the formula $c=round(0.2*P-1)/2$ to calculate the extra sections to overlap. Therefore, when $P=8$ then $c=0$ (equivalent to D), when $P=16$ then $c=1$ (equivalent to O), and so on. Figure \ref{fig:adaptive}

\begin{figure*}
\centering
\includegraphics[width=12cm]{islandNoDisjoint.jpg}
\caption{NOTA: ESTA FIGURA TIENE QUE ACTUALIZARSE CON LA ADAPTATIVA!!!!! Adaptive overlapping algorithm: every island $p$ modifies the  $p+c$,
  $p_{th}$ and $p-c$  components (in grey) of the individuals. $c$ is calculated depending on the number of islands.}
  \label{fig:adaptive}
\end{figure*}
% --------------------------------------------------------------

@@

The four approaches have been run with two different chromosome lengths ($L$): 512 and 2048. Different number of islands ($P$) have also been compared: 8, 16, 32, 64 and 128. This maximum number of islands have also been  used in previous work in the literature \citep{Martens13asynchronous}. The crossover and mutation chosen, SBX and polynomial, have been used previously by other authors in \citep{Durillo08masterslave}.

@@

\begin{table*}
\begin{center}
\begin{tabular}{|c|c|}
\hline
{\em Parameter Name} & {\em Value} \\ \hline
Number of islands ($P$) & 8, 16, 32, 64 and 128 \\ \hline
Chromosome size ($L$) & 512 and 2048 \\ \hline
Execution time (s) & 25 (for 512) and 100 (for 2048) \\ \hline \hline
Global population size ($N$) & 1024 \\ \hline
Selection type & Binary Tournament Selection \\ \hline
Replacement type & Generational \\ \hline 
Crossover type & SBX \\ \hline
Mutation  type & Polynomial\\ \hline
Mutation probability & 1/$L$ \\ \hline
Generations between migration & 5 \\ \hline
Individuals per migration & 1 \\ \hline
Selection for migration & Binary Tournament\\ \hline
Runs per configuration & 30 \\ \hline
\end{tabular}
\caption{Parameters and operators used in the experiments.}
\label{tab:parameters}
\end{center}
\end{table*}

The ECJ framework \citep{ECJ} has been used to run the
experiments. Specific operators have been developed as new modules for
ECJ, and they can be downloaded from our GitHub repository under a
LGPL V3 License
\footnote{\url{https://https://github.com/hpmoon/hpmoon-islands}}. The
island model has been executed synchronously, using the ECJ Internal
Island Model, in a CentOS 5.4 machine with Intel(R) Xeon(R) CPU E5520
@2.27GHz, 16 GB RAM, with Java Version 1.6.0\_16.

@@

Finally, as happened in \citep{Garcia16hpmoon}, results show that all the quality indicators lowers when the number of islands is increased, as the subpopulations are smaller. This is consistent with the claim by \cite{Dorronsoro13superlinear}, who found that cooperative coevolutionary MOEAs work better on bigger populations (more than 100 individuals).

@@

High-performance problems that require a large number of decision variables can leverage the division of the decision space that parallel and distributed algorithms imply. This can be done in dMOEAs by dividing the chromosome into different parts, each one modified by a different island. This paper compares a baseline distributed NSGA-II with three different strategies to separate the chromosome (disjoint or overlapping parts), using a high number of islands. Results show that these methods can achieve better quality metrics than the baseline in the same amount of time, with larger chromosome lengths.

Results also show that when increasing the number of islands, the adaptive overlapping method significantly improves the results with respect to the disjoint and overlapped methods. However, this is only apparent with larger chromosome sizes. Therefore, the length of the chromosome is also a key factor to take into account in adaptive methods, and not only the number of islands. Studying this factor with more types of problems, and new configurations of population size and chromosome lengths will be addressed in the future.%
%

Also, more distributed implementations in several systems (such as a GPUs or heterogeneous clusters) with more different quantities of islands/processors, will be used to perform a scalability study of the different methods, being the transmission time between islands a relevant issue to address. New benchmarks and real problems will also be  used to validate this approach.

@@

TIN2014-56494-C4-3-P y TIN2015-67020-P (Ministerio de Economía y Competitividad), PROY-PP2015-06 (Plan Propio 2015 UGR),  y MSTR (PRY142/14, Fundaci\'on P\'ublica Andaluza Centro de Estudios Andaluces en la IX Convocatoria de Proyectos de Investigaci\'on).
